---
title: "Simulate-data-multi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate-data-multi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(devtools)
#devtools::load_all("~/TedSim/TedSim_1.00/TedSim")
library(TedSim)
```

# Parameter settings
```{r}
ncells <- 1024
phyla <- read.tree(text='((t1:3, (t2:2, t3:2):1):1);')
#phyla <- read.tree(text='((t1:2, t2:2):1, (t3:2, t4:2):1):2;')
N_nodes <- 2*ncells-2
ngenes <- 500
max_walk <- 6
p_a <- 0.6
n_cif <- 30
n_diff <- 20
cif_step <- 0.25
p_d <- 0
N_char <- 32
set.seed(233)
```

#Generate diff-SIFs based on the state tree
```{r}
returnlist <- SIFGenerate(phyla,n_diff,step = cif_step)
sif_mean_raw <- returnlist$sif_mean_raw

sif_label <- sif_mean_raw[[1]][,2]
sif_mean_raw <- lapply(sif_mean_raw,function(X){
  X[,4]
})
sif_mean_raw <- do.call(cbind,sif_mean_raw)
sif_pca <- prcomp(sif_mean_raw)

plot_pca <- data.frame(label=sif_label,x=sif_pca$x[,1],y=sif_pca$x[,2])
p <- ggplot(plot_pca, aes(x, y))
p <- p + geom_point(aes(colour = factor(plot_pca$label)),shape=20) + labs(color='cell states')
p <- p+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
             panel.background = element_blank(), axis.line = element_line(colour = "black"))
p
```

# Generate True Counts and Barcodes
```{r}
#simulate cifs
cifs <- SimulateCIFs(ncells,phyla,p_a = p_a,n_CIF = n_cif,n_diff = n_diff,step = cif_step,p_d = p_d, Sigma = 0.5, N_char = 32, max_walk = max_walk, SIF_res = returnlist, unif_on = FALSE)

#We only need the leaf cells for experiments
cif_leaves <- lapply(c(1:3),function(parami){
  cif_leaves_all <- cifs[[1]][[parami]][c(1:ncells),]
  return(cif_leaves_all)
})
cif_res <- list(cif_leaves,cifs[[2]])
states <- cifs[[2]]
states <- states[1:N_nodes,]
states_leaves <- states[1:ncells,]
muts <- cifs[[7]]
rownames(muts) <- paste("cell",states[,4],sep = "_")
muts_leaves <- muts[1:ncells,]

#simulate true counts
true_counts_res <- CIF2Truecounts(ngenes = 500,ncif = n_cif,ge_prob = 0.3,ncells = ncells, cif_res = cif_res)
```

# Visualize True Counts using tSNE/UMAP
```{r}
tsne_true_counts <- PlotTsne(meta=states_leaves, data=log2(true_counts_res[[1]]+1), cif_type="continuous", n_pc=30, label='cluster', saving = F, plotname="Discrete population (true counts)")
umap_true_counts <- PlotUmap(meta=states_leaves, data=log2(true_counts_res[[1]]+1), n_pc=30, label='cluster', saving = F, plotname="Differentiating population (true counts)")

tsne_true_counts[[2]] + ggtitle("Discrete population (true counts)") + xlab("tSNE 1") + ylab("tSNE 2") +  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30),legend.text = element_text(size = 20))
umap_true_counts[[2]] + ggtitle("Continuous population (true counts)") + xlab("UMAP 1") + ylab("UMAP 2") +  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30),legend.text = element_text(size = 20))
```
# Visualize True Counts with continuous gradient scale
```{r}
library('dichromat')
library(scales)
colors <- hue_pal()(6)
color <- states_leaves[,'cluster']
for (cluster in unique(states_leaves[,'cluster'])){
  depth_range <- sort(unique(states_leaves[states_leaves[,'cluster']==cluster,'depth']))
  edge <- phyla$edge[phyla$edge[,2] == cluster]
  #colfunc<-colorRampPalette(c(colors[edge[1]],colors[edge[2]]))
  colfunc<-colorRampPalette(c('#FFFFFF',colors[edge[2]]))
  color_grad <- colfunc(length(depth_range)+2)
  color_grad <- color_grad[0:-2]
  
  #plot(rep(1,length(depth_range)),col=color_grad,pch=19,cex=3)
  print(color_grad)
  
  color_sub <- states_leaves[states_leaves[,'cluster']==cluster,'depth']
  color_sub[color_sub %in% depth_range] <- color_grad[match(color_sub, depth_range, nomatch = 0)]
  color[color == cluster] <- color_sub
}
states_leaves <- cbind(states_leaves,color)

# create a character vector of colornames
colr <- as.character(unique(states_leaves[,'color']))
plot_umap <- umap_true_counts[[1]] 
p <- ggplot(plot_umap, aes(x, y))
p <- p + geom_point(aes(colour = as.factor(states_leaves[,'color'])),shape=20,size = 5) + labs(color='cell state')
p <- p+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))
p <- p + scale_color_manual(breaks=unique(states_leaves[,'color']), values=colr)
p + theme(legend.position = "none") + ggtitle("Continuous population (true counts)") + xlab("UMAP 1") + ylab("UMAP 2") +  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30),legend.text = element_text(size = 20))
```

# Transform True counts to observed counts, and visualize
```{r}
data(gene_len_pool)
gene_len <- sample(gene_len_pool, ngenes, replace = FALSE)
observed_counts <- True2ObservedCounts(true_counts=true_counts_res[[1]], meta_cell=true_counts_res[[3]], protocol="UMI", alpha_mean=0.2, alpha_sd=0.05, gene_len=gene_len, depth_mean=1e5, depth_sd=3e3)
umap_UMI_counts <- PlotUmap(meta=states_leaves, data=log2(observed_counts[[1]]+1), n_pc=30, label='cluster', saving = F, plotname="Differentiating population (observed counts)")
umap_UMI_counts[[2]] + ggtitle("Continuous population (observed counts)") + xlab("UMAP1") + ylab("UMAP 2") +  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30),legend.text = element_text(size = 20))
```
# Simulate realistic barcode data with similar mutated states as real datasets
```{r}
#visualize simulated mutation distribution
muts_table <- table(unlist(muts))
muts_table <- muts_table[dimnames(muts_table)[[1]]!='0']
muts_table <- muts_table[dimnames(muts_table)[[1]]!='-']
muts_prop <- prop.table(sort(muts_table,decreasing = TRUE))
muts_prop <- muts_prop[1:15]
df_prop <- as.data.frame(muts_prop)
p <- ggplot(df_prop) + geom_col(aes(x = Var1,y = Freq))
p <- p + ggtitle("Distribution of mutations: Fitted data") +
  xlab("Mutated States") + ylab("Freq")
p + theme(title =element_text(size=20),axis.text = element_text(size = 18), axis.title = element_text(size = 25))
#ggsave('~/TedSim/tedsim_plots_paper/dist_mutation_fitted_embryo2.png')
```
# Generate capture dropout
```{r}
muts <- CaptureDrop(observed_counts[[1]],muts)
```


# Generate combined profiles of lineage barcodes and gene expressions
```{r}
profile_res <- Generate_profile_multi(observed_counts,muts,states_leaves)
profile_out <- profile_res[[1]]
top_genes <- profile_res[[2]]
```

#Output files
```{r}
gene_expression_dir <- "./output/counts_tedsim_pa0.75_step0.25.csv"
cell_meta_dir <- "./output/cell_meta_tedsim_pa0.75_step0.25.csv"
character_matrix_dir <- "./output/character_matrix_pd0.txt"
combined_profile_dir <- "./output/profile_c_tedsim.txt"
top_genes_dir <- "./output/top_genes_tedsim.txt"
tree_gt_dir <- "./output/tree_gt_bin_tedsim_pd0.newick"

write.tree(cifs[[4]],tree_gt_dir)
write.csv(observed_counts[[1]],gene_expression_dir,row.names = FALSE)
write.csv(states_leaves,cell_meta_dir)
write.table(muts_leaves,character_matrix_dir)
write.table(profile_out,file = combined_profile_dir,row.names = FALSE,sep = "\t", quote = FALSE)
write.table(subset(top_genes,select = 1),top_genes_dir,row.names = FALSE,sep = "\t", quote = FALSE)


```

